---
title: "DMOZ"
author: "Kenneth E. Shirley"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{DMOZ}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r setup, echo = FALSE}
knitr::opts_chunk$set(collapse = FALSE, comment = "#>", cache = TRUE)
```

Here we'll describe how to compute maximum entropy summary trees to visualize the topic hierarcy of the URLs in DMOZ (aka the Open Directory Project).

First, we downloaded the DMOZ "content" data in April of 2015 from [http://www.dmoz.org/rdf.html](http://www.dmoz.org/rdf.html). This file was about 1.8 GB unzipped.

Next, we use the command line to extract just the assigned topic of each of the approximately 3.77 million URLs and write them, one per line, to a file called "dmoz-topics.txt".

    grep "<topic>" content.rdf.u8 | sed 's/<topic>//g' | sed 's/<\/topic>//g' | sed 's/^    //g' > dmoz-topics.txt

Next, we read this list of topics into R and compute their frequencies:
``` {r, dmoz-read}
raw <- readLines("~/Stats/Trees/dmoz/dmoz-topics.txt")

# compute the frequency of each unique topic:
dmoz.table <- table(raw)

# order them alphabetically
dmoz.table <- dmoz.table[order(names(dmoz.table))]

# separate into names + weights:
full.node <- names(dmoz.table)  # full node names with slashes
weight <- as.integer(dmoz.table)  # weights
```

Now we have to assign each node an ID number and compute the ID number of the node's parent. The only complication here is that some nodes do not currently have their parent node listed in the data, because the parent node is an internal node to which zero URLs are directly assigned.

The following block of code starts with the nodes of the tree that have nonzero weights, looks for their parents, and if any of their parents are not in the tree already, they are added to the tree.

```{r fill-internal-nodes}
# initialize variables
parent <- NULL  # the vector of all parents
n.added <- NULL  # how many new internal nodes added each iteration
iter <- 0  # count iterations (just out of curiosity)
to.split <- full.node  # the set of full names to split
new.parent <- rep(NA, length(weight))  # parents of each round of nodes

t1 <- Sys.time()
while (sum(is.na(new.parent)) > 0) {
  iter <- iter + 1
  #print(iter)

  # split by slash and extract the leaf label of each node, and the 'stem'
  dmoz.split <- strsplit(to.split, "/")
  label <- sapply(dmoz.split, function(x) tail(x, 1))
  stem <- sapply(dmoz.split, function(x) paste(x[-length(x)], collapse = "/"))

  # compute the parent of each node:
  new.parent <- match(stem, full.node)

  # if new.parent is NA, then we have to add an internal node
  # get unique internal nodes that must be added
  new.internal.nodes <- unique(stem[is.na(new.parent)])
  n.added <- c(n.added, length(new.internal.nodes))
  # add the new internal nodes to the full list
  full.node <- c(full.node, new.internal.nodes)
  # internal nodes have a weight of zero by definition here
  weight <- c(weight, rep(0, length(new.internal.nodes)))
  # set up the next set of nodes whose parents must be found
  to.split <- new.internal.nodes
  # add to the vector of parents
  parent <- c(parent, match(stem, full.node))
}
t2 <- Sys.time()
t2 - t1
```


Now we compute the labels, we assemble the nodes, parents, weights, and labels in a data frame, and we clean up one pseudo-node that was computed as the parent of the root.

``` {r assemble}
label <- sapply(strsplit(full.node, "/"), function(x) tail(x, 1))

# There should be one that is the 'parent' of the root, which is an empty node
# Give it a label of NA
label[sapply(label, length) == 0] <- NA
label <- unlist(label)

# Pull it all into a data.frame:
dmoz <- data.frame(node = 1:length(full.node),
                   parent = parent,
                   weight = weight,
                   label = label,
                   stringsAsFactors = FALSE)

# identify the 'parent' of the root, which doesn't really exist:
to.remove <- which(is.na(dmoz[, "label"]))

# Set the parent of the root to zero
dmoz[dmoz[, "parent"] == to.remove, "parent"] <- 0

# remove the 'parent' of the root
dmoz <- dmoz[-to.remove, ]

# Look at the data a bit:
dim(dmoz)
dmoz[1:10, ]
table(dmoz[, "weight"] > 0)
```

Now we load the `summarytrees` package and compute maximum entropy summary trees using the greedy algorithm for this data for $k = 1, 2, ..., K = 100$.

``` {r greedy}
library(devtools)
install_github("kshirley/summarytrees")
library(summarytrees)

# compute a set of K summary trees:
t1 <- Sys.time()
K <- 100
g <- greedy(node = dmoz[, "node"],
            parent = dmoz[, "parent"],
            weight = dmoz[, "weight"],
            label = dmoz[, "label"],
            K = K)
t2 <- Sys.time()
t2 - t1

# Prepare the summary trees for the visualization:
json <- prepare.vis(tree.list = g$summary.trees,
                    labels = g$data[, "label"],
                    tree = g$tree,
                    legend.width = 150,
                    node.width = 225,
                    node.height = 12,
                    units = "# of URLs",
                    print.weights = TRUE,
                    legend.color = "lightsteelblue",
                    color.level = 2)
```

Last, serve the vis in the browser:
``` {r serve, eval = FALSE}
draw.vis(json.object = json,
         out.dir = tempfile(),
         open.browser = interactive())
```

To see this set of 100 maximum entropy summary trees, go [here](http://www.research.att.com/~kshirley/summarytrees/dmoz).

From the visualization, we can see that:

1. The biggest second-level topic is "World", which branches out into lots of different individual languages, the biggest of which is German (in terms of the number of URLs in the directory assigned there.)

2. The second-level topic "Regional" is comprised mostly of European locations and North American ones. The United States subtree branches out into individual states.

3. The largest second-level topics besides "World" and "Regional" are "Business", "Arts", and "Society". Despite having more descendants than "Arts" and "Society", the children of "Business" only appear in the summary trees when $k = 99$, after some children of the other two nodes have appeared. This is because the children of business are relatively small and evenly distributed, thus they didn't increase the overall entropy very much until several of them could be expanded at the same time.























