---
title: "Computing and Visualizing Summary Trees"
author: "Kenneth E. Shirley"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Computing and Visualizing Summary Trees}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

This vignette demonstrates how to compute and visualize summary trees. We will do this using two data sets that can be represented by rooted, node-weighted trees, where the node weights are non-negative:

1. The flare software package, where the nodes of the tree are files or directories, the weights are file sizes in bytes (where directories have size zero), and the tree structure is given by the file structure of the package directory.

2. The webpage categories from the Open Directory Project, also known as DMOZ. Here, the nodes of the tree are categories of webpage types, the weights are the counts of how many webpages in DMOZ belong to each category, and the tree structure is given by the category hierarchy created by the DMOZ editors.

First, load the package:
```{r}
library(summarytrees)
```

## Data
The raw 'flare' data was downloaded as a .json file from Mike Bostock's gist located here:

https://gist.github.com/mbostock/1093025

Next, we converted the json file to a data.frame in R using the following code:

[insert code here]

This data.frame version of the data is included in the `data` directory of the `summarytrees` package. To load it, just type:

```{r}
data(flare)
dim(flare)
head(flare)
```

## Styles

The `html_vignette` template includes a basic CSS theme. To override this theme you can specify your own CSS in the document metadata as follows:

    output: 
      rmarkdown::html_vignette:
        css: mystyles.css

## Figures

The figure sizes have been customised so that you can easily put two images side-by-side. 

```{r, fig.show='hold'}
plot(1:10)
plot(10:1)
```

You can enable figure captions by `fig_caption: yes` in YAML:

    output:
      rmarkdown::html_vignette:
        fig_caption: yes

Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**.

## More Examples

You can write math expressions, e.g. $Y = X\beta + \epsilon$, footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`.

```{r, echo=FALSE, results='asis'}
knitr::kable(head(mtcars, 10))
```

Also a quote using `>`:

> "He who gives up [code] safety for [code] speed deserves neither."
([via](https://twitter.com/hadleywickham/status/504368538874703872))
